### 2022-07-18 paper/localpsf.tex
tightening up paper. commenting out stuff and reducing focus on H-matrices

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. H-matrix approximations are highly desirable because there are fast H-matrix methods to add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known, which allows us to construct the H-matrix. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator applies.
\end{abstract}

@@

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High-rank matrix free approximation methods are needed.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.4]{bct1.eps}
	\end{center}
	\caption{Hierarchical matrix block structure for the Stokes Hessian. The red blocks (tiny) are dense. All other blocks are low rank.
	}
	\label{fig:hmatrix_intro}
\end{figure}

Our method forms a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of the domain. 
	}
	\label{fig:batches_intro}
\end{figure}

Classical H-matrix construction methods (described in Appendix \ref{sec:H_matrix_construction}) require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free. Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}, which we do not use here. 
These alternative methods have, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
While methods based on the peeling process are asymptotically scalable in theory, in practice the required number of matrix vector products is large. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

We construct the H-matrix using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate shifted versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

Picking the batches of points $x_i$ requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

@@

@@

Let $C$, $L^i$, and $Q^{ij}$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) := x^i x^j
\end{equation*}
for $i,j=1,\dots,\gdim$. Using the definition of $\spatialvol$ in \eqref{eq:define_vol} and using \eqref{eq:operator_simultaneous_ips}, we have
\begin{equation*}
	\spatialvol(x) = \int_\Omega \phi_x(y) C(y) ~ dy = \left(\phi_x, C\right)_{L^2(\Omega)} = \left(\mathcal{A}^T C\right)^*(x).
\end{equation*}
Hence, we may compute $\spatialvol(x)$ for all $x$ simultaneously by applying $\Aop^T$ to $C$. Analogous manipulations show that $\spatialmean(x)$ and $\spatialcov(x)$ may be computed for all points $x$ simultaneously by applying $\Aop^T$ to the functions $L^i$ and $Q^{ij}$, respectively. To summarize the results of doing this, we have
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i,j=1,\dots, \gdim$. Here $f/g$ denotes pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denotes pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The resulting method for computing discretized versions of $\spatialvol$, $\spatialmean$, and $\spatialcov$ is shown in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	{	
		\Input{Operator $\bm{A}^T$}
		\Output{$\bm{\spatialvol}\in \mathbb{R}^\fedim$, $\bm{\spatialmean}\in \mathbb{R}^{\gdim \times \fedim}$, and $\bm{\spatialcov} \in \mathbb{R}^{\gdim \times \gdim \times \fedim}$}

		Form vector $\bm{C} \in \mathbb{R}^\fedim_{\massmatrix}$ by either projecting or interpolating constant function $C(x)=1$ onto $V_h$

		$\bm{\spatialvol} \gets \massmatrix^{-1}\bm{A}^T \bm{C}$

		\For{$i=1,2,\dots,\gdim$}{
			Form vector $\bm{L}^i \in \mathbb{R}^\fedim_{\massmatrix}$ by either projecting or interpolating linear function $L^i(x) = x^i$ onto $V_h$

			$\bm{\spatialmean}^i \gets \left(\massmatrix^{-1}\bm{A}^T \bm{L}^i\right) / \bm{\spatialvol}$
		}
		\For{$i=1,2,\dots,\gdim$}{
			\For{$j=1,\dots,i$}{
				Form quadratic function $\bm{Q}^{ij} = x^i x^j$

				$\bm{\spatialcov}^{ij} \gets \left(\massmatrix^{-1}\bm{A}^T \bm{Q}^{ij}\right) / \bm{\spatialvol} - \bm{\spatialmean}^i\cdot \bm{\spatialmean}^j$

				$\bm{\spatialcov}^{ji} \gets \bm{\spatialcov}^{ij}$

			}
		}

	}
	\caption{Compute discretized scaling factor $\mathbf{\spatialvol}$, mean $\boldsymbol{\spatialmean}$, and covariance $\boldsymbol{\spatialcov}$}
	\label{alg:varhpi_mean_cov}
\end{algorithm2e}

@@

When $\Aop$ is symmetric positive semi-definite, $\mathbf{A}_H$ may be non-symmetric and indefinite due to errors in the approximation. In this case, one may (optionally) modify the H-matrix representation of $\mathbf{A}_H$ to make it symmetric and nearly positive semi-definite. The matrix is symmetrized by computing $\left(\mathbf{A}_H+\mathbf{A}_H^T\right)/2$ using H-matrix addition. Negative eigenvalues that are large or medium in magnitude are computed, then flipped in sign to become positive. Small negative eigenvalues may remain because they are difficult to compute numerically. This is described in Appendix \ref{sec:make_spd}.

@@

		Compute $\bm{\spatialvol}$, $\bm{\spatialmean}$, and $\bm{\spatialcov}$ using Algorithm \ref{alg:varhpi_mean_cov}.

@@

We form our H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming H-matrix representations $\mathbf{\Aker}_H$ and $\massmatrix_H$ of $\mathbf{\Aker}$ and $\massmatrix$, respectively, then using fast H-matrix methods to multiply these matrices per \eqref{eq:boldA} to form
\begin{equation*}
\mathbf{A}_H = \massmatrix_H \mathbf{\Aker}_H \massmatrix_H.
\end{equation*}
Once $\mathbf{A}_H$ is formed, useful matrix operations such as matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, and matrix inversion may be performed using fast scalable H-matrix methods \cite{BORMHACKBUSCHBOOK}. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead. For more details on H-matrices, see \cite{Hackbusch15}.

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{Kriemann08}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank^2 \fedim \log \fedim)$ entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(p_i, p_j)$, and these entries are computed via the procedure described in Section \ref{sec:approximate_kernel_entries}. Here $\hrank$ is the rank of the highest-rank block in the H-matrix. A dense representation of $\mathbf{\Aker}$ is never formed. We describe this H-matrix construction process in detail in Appendix \ref{app:h_matrix}. We form $\massmatrix_H$ using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same recursive block partitioning structure as was used for $\mathbf{\Aker}$.

@@

\paragraph{(2) Building the H-matrix} The classical H-matrix construction technique, which is described in Appendix \ref{app:h_matrix}, requires evaluating $O(\hrank^2 \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulses used in the RBF interpolation. This is done using a precomputed KD tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, we must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, we must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations, which typically is negligible. Therefore, building the H-matrix requires
\begin{equation*}
	O\left(\left(\hrank^2 \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well-known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion, and low rank updates require
\begin{equation}
\label{eq:H_matrix_cost_operations}
	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
\end{equation}
where $a \in \{0,1,2,3\}$ \cite{GrasedyckHackbusch03}. The precise cost depends on the type of H-matrix used and the operation being performed. In this paper, we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian. Then, we use one matrix inversion to form the inverse of the Hessian, which is used as a preconditioner for the Newton linear system. We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix \ref{sec:recycle_krylov}. The rational method for modifying $\mathbf{A}_H$ to be positive semi-definite, described in Appendix \ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the order of the rational function used.

@@

\section{Hierarchical matrix details}
\label{app:h_matrix}

Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank. Such matrices are known as hierarchical matrices (H-matrices). Many classes of H-matrices exist (H1, H2, HSS, HBS, and more), and all of these types of H-matrices could be used in conjunction with our product-convolution approximation. Here we use classical H1 matrices. For this section, when we say H-matrix, we are referring to H1 matrices.

>>>>

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for approximating operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. 
%	Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. 
	Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Fast H-matrix methods are used to perform further matrix computations including matrix-matrix addition and multiplication, matrix inversion and factorization, and matrix-vector products in nearly-linear complexity.
	We apply the method to approximate the Hessian in a large-scale PDE-constrained inverse problem with highly informative data. Numerical results demonstrate that our method substantially outperforms existing methods. We are able to approximate high rank operators using only a small number of operator applies.
\end{abstract}

@@

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High-rank approximation methods are needed.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{impulse_batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{impulse_batch2.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{impulse_batch3.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. 
%	Brightness scale differs between the subfigures. 
	Black stars indicate point source locations. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of the domain. 
	}
	\label{fig:batches_intro}
\end{figure}

The central challenge we face is that existing scalable high-rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

Picking the batches of points requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

@@

%\begin{figure}
%	\begin{center}
%		\includegraphics[scale=0.4]{bct1.eps}
%	\end{center}
%	\caption{Hierarchical matrix block structure for the Stokes Hessian. The red blocks (tiny) are dense. All other blocks are low rank.
%	}
%	\label{fig:hmatrix_intro}
%\end{figure}

The ability to rapidly evaluate approximations of arbitrary entries of $\Aop$'s integral kernel allows us to form a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank. 
%(see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix is scales as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear} in $N$. The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

%Classical H-matrix construction methods 
%(described in Appendix \ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.
Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}, which we do not use here. 
These alternative methods have been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
Methods based on the peeling process are asymptotically scalable (typically the number of matrix vector products required scales as $O(\log N)$), but in practice the required number of matrix vector products is large. Improving the peeling process to reduce its cost is currently an area of active research \cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently a topic of considerable research, so more efficient matrix-free H-matrix methods based on the peeling process may be discovered in the future.

@@

Let $C$, $L^i$, and $Q^{ij}$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) := x^i x^j
\end{equation*}
for $i,j=1,\dots,\gdim$. Using the definition of $\spatialvol$ in \eqref{eq:define_vol} and using \eqref{eq:operator_simultaneous_ips}, we have
\begin{equation*}
	\spatialvol(x) = \int_\Omega \phi_x(y) C(y) ~ dy = \left(\phi_x, C\right)_{L^2(\Omega)} = \left(\mathcal{A}^T C\right)^*(x).
\end{equation*}
Hence, we may compute $\spatialvol(x)$ for all $x$ simultaneously by applying $\Aop^T$ to $C$. Analogous manipulations show that $\spatialmean(x)$ and $\spatialcov(x)$ may be computed for all points $x$ simultaneously by applying $\Aop^T$ to the functions $L^i$ and $Q^{ij}$, respectively. To summarize the results of doing this, we have
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i,j=1,\dots, \gdim$. Here $f/g$ denotes pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denotes pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$.
%The resulting method for computing discretized versions of $\spatialvol$, $\spatialmean$, and $\spatialcov$ is shown in Algorithm \ref{alg:varhpi_mean_cov}.

%\begin{algorithm2e}
%	\SetAlgoNoLine
%	\SetKwInOut{Input}{Input}
%	\SetKwInOut{Output}{Output}
%	{	
%		\Input{Operator $\bm{A}^T$}
%		\Output{$\bm{\spatialvol}\in \mathbb{R}^\fedim$, $\bm{\spatialmean}\in \mathbb{R}^{\gdim \times \fedim}$, and $\bm{\spatialcov} \in \mathbb{R}^{\gdim \times \gdim \times \fedim}$}
%		
%		Form vector $\bm{C} \in \mathbb{R}^\fedim_{\massmatrix}$ by either projecting or interpolating constant function $C(x)=1$ onto $V_h$
%		
%		$\bm{\spatialvol} \gets \massmatrix^{-1}\bm{A}^T \bm{C}$
%		
%		\For{$i=1,2,\dots,\gdim$}{
%			Form vector $\bm{L}^i \in \mathbb{R}^\fedim_{\massmatrix}$ by either projecting or interpolating linear function $L^i(x) = x^i$ onto $V_h$
%			
%			$\bm{\spatialmean}^i \gets \left(\massmatrix^{-1}\bm{A}^T \bm{L}^i\right) / \bm{\spatialvol}$
%		}
%		\For{$i=1,2,\dots,\gdim$}{
%			\For{$j=1,\dots,i$}{
%				Form quadratic function $\bm{Q}^{ij} = x^i x^j$
%				
%				$\bm{\spatialcov}^{ij} \gets \left(\massmatrix^{-1}\bm{A}^T \bm{Q}^{ij}\right) / \bm{\spatialvol} - \bm{\spatialmean}^i\cdot \bm{\spatialmean}^j$
%				
%				$\bm{\spatialcov}^{ji} \gets \bm{\spatialcov}^{ij}$
%				
%			}
%		}
%		
%	}
%	\caption{Compute discretized scaling factor $\mathbf{\spatialvol}$, mean $\boldsymbol{\spatialmean}$, and covariance $\boldsymbol{\spatialcov}$}
%	\label{alg:varhpi_mean_cov}
%\end{algorithm2e}

@@

When $\Aop$ is symmetric positive semi-definite, $\mathbf{A}_H$ may be non-symmetric and indefinite due to errors in the approximation. In this case, one may (optionally) modify the H-matrix representation of $\mathbf{A}_H$ to make it symmetric positive semi-definite using the rational method described in Appendix \ref{sec:make_spd}.

%. The matrix is symmetrized by computing $\left(\mathbf{A}_H+\mathbf{A}_H^T\right)/2$ using H-matrix addition. Negative eigenvalues modified to be positive using a rational method that is described in Appendix \ref{sec:make_spd}.

@@

		Compute $V, \mu$, and $\Sigma$ (Section \ref{sec:intromoments})

%		Compute $\bm{\spatialvol}$, $\bm{\spatialmean}$, and $\bm{\spatialcov}$ using Algorithm \ref{alg:varhpi_mean_cov}.

@@

We form our H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming H-matrix representations $\mathbf{\Aker}_H$ and $\massmatrix_H$ of $\mathbf{\Aker}$ and $\massmatrix$, respectively, then using fast H-matrix methods to multiply these matrices per \eqref{eq:boldA} to form
\begin{equation*}
\mathbf{A}_H = \massmatrix_H \mathbf{\Aker}_H \massmatrix_H.
\end{equation*}
%Once $\mathbf{A}_H$ is formed, useful matrix operations such as matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, and matrix inversion may be performed using fast scalable H-matrix methods \cite{BORMHACKBUSCHBOOK}. 
We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead. For more details on H-matrices, see \cite{Hackbusch15}.

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{Kriemann08}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank^2 \fedim \log \fedim)$ entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(p_i, p_j)$, and these entries are computed via the radial basis function interpolation method described in Section \ref{sec:approximate_kernel_entries}. Here $\hrank$ is the rank of the highest-rank block in the H-matrix. 
A dense representation of $\mathbf{\Aker}$ is never formed. 
%We describe this H-matrix construction process in detail in Appendix \ref{app:h_matrix}. 
We form $\massmatrix_H$ using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same recursive block partitioning structure as was used for $\mathbf{\Aker}$.

@@

\paragraph{(2) Building the H-matrix} The classical H-matrix construction technique
%, which is described in Appendix \ref{app:h_matrix}, 
requires evaluating $O(\hrank^2 \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulses used in the RBF interpolation. This is done using a precomputed KD tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, we must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, we must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations, which typically is negligible. Therefore, building the H-matrix requires
\begin{equation*}
	O\left(\left(\hrank^2 \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well-known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion, and low rank updates require
\begin{equation}
\label{eq:H_matrix_cost_operations}
	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
\end{equation}
where $a \in \{0,1,2,3\}$ \cite{GrasedyckHackbusch03}. The precise cost depends on the type of H-matrix used and the operation being performed. In this paper, we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian. Then, we use one matrix inversion to form the inverse of the Hessian, which is used as a preconditioner for the Newton linear system. 
%We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix \ref{sec:recycle_krylov}.
The rational method for modifying $\mathbf{A}_H$ to be positive semi-definite, described in Appendix \ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the order of the rational function used.

@@

%\section{Hierarchical matrix details}
%\label{app:h_matrix}
%
%Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank. Such matrices are known as hierarchical matrices (H-matrices). Many classes of H-matrices exist (H1, H2, HSS, HBS, and more), and all of these types of H-matrices could be used in conjunction with our product-convolution approximation. Here we use classical H1 matrices. For this section, when we say H-matrix, we are referring to H1 matrices. 
%
%While a dense $\fedim \times \fedim$ matrix traditionally requires $O(\fedim^2)$ memory to store, H-matrices may be stored using $O(\fedim \log \fedim)$ memory, by storing only the low rank factors for the low rank blocks. Recursive algorithms can take advantage of the H-matrix low rank block structure to perform matrix arithmetic fast. Conventional dense matrix algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix-matrix addition require either $O(\fedim^2)$ or $O(\fedim^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(\fedim \log(\fedim)^a)$ time and memory for H-matrices. Here $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix. For more details on H-matrices, we recommend \cite{HMATRIXGOOD}. 
%
%
%\subsection{H-matrix construction}
%\label{sec:H_matrix_construction}
%
